---
title: 'DATA 607: Week 2'
author: "Cindy Lin"
date: "2025-2-5"
output: html_document
---


# INTRODUCTION
The purpose is to survey at least 5 people on their rating of the latest popular movies. Part I and II were done by building a table in MySQLWorkbench.

## Installing and Loading library

```{r library, message= FALSE}
install.packages('RMySQL', repos='http://cran.us.r-project.org')
library(RMySQL) 

```
I am installing and loading the necessary package and library to load the information from SQL database to R. 

## Transfer data from SQL database
```{r data transfer, message=FALSE}

username <- Sys.getenv("USERNAME")
password <- Sys.getenv("password")

database = dbConnect(RMySQL::MySQL(), user= username, password= password, dbname='cindy.lin79', port=3306, host='cuny607sql.mysql.database.azure.com')

data <- dbGetQuery(database, 'SELECT * FROM `cindy.lin79`.top_popular_movies')

print(data)
```
This is Part 3 of the assignment where "Load the information from the SQL database into an R dataframe."

## Part 4 and Data Visualization 
```{r omitting nulls}
library(ggplot2)
library(dplyr)

data_no_data <- na.omit(data)
## This removes all the null data which is the approach I decided to use to handle nulls

Average_rating <- data_no_data %>%
  group_by (movie_names) %>%
  summarize (
    avg_rating = mean(rating, na.rm = TRUE), count = n()
  )


print(Average_rating)
  
ggplot(data=data_no_data, aes(x = movie_names, y = rating)) + geom_boxplot()

```


I removed the null data from the data set which removed two movies. The movies that were excluded from the analysis are: Wicked and Dune: Part Two. 


# CONCLUSION

There were two movies that all respondents did not see and I decided to exclude them. The objective is to have the ratings of movies and if there are no ratings, it's in a separate category which is "Not Seen". We can see in the box plot that the ratings for "Inside Out 2" and "Venom: The Last Dance" did not have any variations, which can be attributed to only a few respondents who haven't seen the movie. 

As a recommendation, I would increase the surveyed population to lessen the possibility that we have to exclude movies because no one has seen them. 

