---
title: 'DATA 607: Week 5'
author: "Cindy Lin"
date: "2025-02-24"
output:
  pdf_document: default
  html_document: default
  
---

# INTRODUCTION

Week 5 project's goal is to transform a text file of a certain structure to a CSV file that display the following information:

Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents


## Loading library

```{r library, include= FALSE}
library (tidyverse)

```
I am loading the tidyverse library because there are functions that can help tidy the loaded data

## Loading the data and removing rows
```{r getdata, message=FALSE}

get_data <- read.delim("tournamentinfo.txt", sep = "|", header = TRUE, skip = 1)
# Skip the first row to get the header

get_data <- get_data[!grepl("--------", get_data$Pair), ] 
# Remove "-----"

get_data <- get_data[-1,]
# Remove the first line

glimpse(get_data)

```
Load the text file from my working directory and removing non-data rows.

## Changing the table structures 

```{r table structure}
odd_rows <- get_data[seq(1, nrow(get_data), by = 2), , drop = FALSE]
#This had the player name, number of points and the rounds

odd_rows <- odd_rows[grepl("\\d", odd_rows$Pair), ]
#remove empty rows

odd_rows <- odd_rows %>% 
  mutate_at(vars(4:10), ~ str_replace_all(., "[^0-9]", ""))
#remove the letter in the round columns

even_rows <- get_data[seq(2, nrow(get_data), by = 2), , drop = FALSE]
#This has the state and the pre-rating 

even_rows <- even_rows %>% mutate(extracted_number = 
             str_extract(Player.Name, "(?<=R:).*?(?=->)"))
#extract the player's pre-rating

even_rows <- even_rows %>% 
  mutate(extracted_digits = ifelse 
  (str_detect(trimws(even_rows$extracted_number), "P"), 
   str_extract(trimws(even_rows$extracted_number), "^\\d+(?=P)"), 
   trimws(even_rows$extracted_number)))
#remove the "P" in some of the pre-rating

even_rows <- even_rows %>% mutate(Ranking = as.numeric(even_rows$extracted_digits))
#make the ranking to be numeric

df_restructured <- data.frame(Odd_Values = odd_rows, 
                             Even_Values = even_rows$Pair, even_rows$Ranking)
#combine the two tables (even and odd)

df_col_rename <- df_restructured %>% rename(
  Pair = Odd_Values.Pair,
  Player_name = Odd_Values.Player.Name,
  Total_Num_of_Points = Odd_Values.Total,
  Round_1 = Odd_Values.Round,
  Round_2 = Odd_Values.Round.1,
  Round_3 = Odd_Values.Round.2,
  Round_4 = Odd_Values.Round.3,
  Round_5 = Odd_Values.Round.4,
  Round_6 = Odd_Values.Round.5,
  Round_7 = Odd_Values.Round.6,
  Player_State = Even_Values,
  Player_PreRating = even_rows.Ranking
  # renamed all the columns in preparation for the final result
  
)

```
Since the original data set had data in different location (where part of the data are in the odd rows and the other part are in even rows), I need to organize it in a way that I can transform it to the final format where all the pieces is in one place. So I separate the pieces in a consistent format and then re-combine them again that results in all the data in their respective columns.  

## Transform wide to long
``` {r long format}

Long_format <- df_col_rename%>%
  pivot_longer(cols = contains("Round"), names_to = "Round", values_to = "Opponent_ID") %>%
  drop_na(Opponent_ID)
# Transform to long format where all the round columns (1,2,3..) is now just one column 
#called Round and the value in those columns are now in new column called Opponent ID

Long_format <- Long_format %>%
  mutate(Opponent_ID = as.numeric(Opponent_ID))

df_col_rename <- df_col_rename %>%
  mutate(Pair = as.numeric(Pair))
#prepare the data type to be the same data type for joining


```
I need to get the average as part of the final result so I transform the wide structure to a long format to prepare to aggregate later. 

## Joining tables 
```{r joining table }
df_with_opponent_rank <- Long_format %>%
  left_join(df_col_rename %>% select(Pair, Player_PreRating), by = c("Opponent_ID" = "Pair"))
#joined Long_format and df_col_rename to get the opponent rank

```
Since I need to get the ranking of the opponents, I created another column that match the opponent's ID and pull the ranking next to that column. This way I have the player and a list of their opponents and their opponent's ranking.

## Averaging the Opponent's Pre-Rating
```{r }
final_table <- df_with_opponent_rank %>%
  group_by(Pair,Player_name, Player_State, Total_Num_of_Points, Player_PreRating.x) %>%
  summarise(Avg_Pre_Chess_Rating_of_Opponents = round(mean(Player_PreRating.y, na.rm = TRUE)))
            
```
I averaged the opponent's pre-rating and grouped it by the rest of the attributes which ends up giving me a player per row. 

## Generate CSV file
```{r }
write.csv(final_table, "tournamentinfo.csv")

```
Saved the CSV file to working directory

# CONCLUSION
For this project, I was able to utilize the skills learned from prior weeks. In particular, loading and transforming data (wide to long) and regular expressions (retaining digits, dropping letters, etc).

Take Away from the Assignment-  
*Sequencing (seq) rows - allows for getting even or odd rows which is another tool to help transform the data in the desire format. 

*Joining table - similarity to SQL (which allows for the search and match type of function I needed for averaging the opponent's ranking)