---
title: 'DATA 607: Week 6'
author: "Cindy Lin and William Forero"
date: "2025-03-05"
output:
  pdf_document: default
  html_document: default
  
---

# INTRODUCTION

Week 6 project’s goal is to transform three untidy data set and tidy/transform them and generate the
discussed analysis.
Data Set#1 - tax dataset for 2022

## Loading library

```{r library}

library (tidyverse)
library (janitor)

```
I am loading the tidyverse library because there are functions that can help tidy the loaded data. Janitor library is needed for the clean_names function.

## Loading the data and removing rows
```{r getdata, message=FALSE}
get_data <- read.csv("22.csv", skip = 3)
# Skip the first row to get the header
get_data <- get_data %>% select(-c(22:25))
# Remove duplicate entries
glimpse(get_data)

```
Loading the data 


## Renaming columns
```{r columns}

colnames(get_data) <- c(
"Adjusted.gross.income",
"All_Returns_Count",
"All_Returns_Percent",
"All_Adjusted.gross.income.less.deficit_Amount",
"All_Adjusted.gross.income.less.deficit_Percent",
"All_Average.adjusted.gross.income.less.deficit_Amount",
"Taxable_Returns_Amount",
"Taxable_Returns_Percent",
"Taxable_Adjusted.gross.income.less.deficit_Amount",
"Taxable_Adjusted.gross.income.less.deficit_Percent",
"Taxable_Taxable.Income_Count",
"Taxable_Taxable.Income_Amount",
"Taxable_Taxable.Income_Percent",
"Taxable_Income.tax.after.credits_Count",
"Taxable_Income.tax.after.credits_Amount",
"Taxable_Income.tax.after.credits_Percent",
"Taxable_Total.Income.Tax_Amount",
"Taxable_Total_Percent",
"Taxable_Total_Taxable.income_Percent",
"Taxable_Total_Adjusted.gross.income.less.deficit_Percent",
"Taxable_Average.total.income.tax_Amount"
)
# renamed the columns
df <- get_data[-c(27:198), ]
# removed empty rows
df <- df[-c(1:7), ]
#removed row 1 - 7 (not needed information)
df <- df %>% clean_names()
#clean the column names to make it consistent
glimpse(df)
```
Renamed the columns as this is important for the next step when I have to format the structure to a long
format. The column names will point to how othe structure will be. For example, the period and underscore
of the naming convention is intentional as the underscore serves as the separator.

# Long format
```{r long format}

df_tidy <- df %>%
pivot_longer(
cols = -`adjusted_gross_income`,
names_to = c("Return_Type", "Category", "Metric"),
names_pattern = "^([^_]+)_(.*)_([^_]+)$",
values_to = "Value"
)
glimpse(df_tidy)
```
Shaping to long format for analysis

# Total tax returns filed
```{r }
df_filter <- df_tidy %>%
filter(Return_Type == "all", Category == "returns", Metric == "count")
# filtering taxable returns and the percentage for ggplot
df_filter <- df_filter %>%
mutate(adjusted_gross_income = fct_collapse(adjusted_gross_income,
"<25K" = c("No adjusted gross income",
"$1 under $5,000",
"$5,000 under $10,000",
"$10,000 under $15,000",
"$15,000 under $20,000",
"$20,000 under $25,000"),
"25K - <75K" = c("$25,000 under $30,000",
"$30,000 under $40,000",
"$40,000 under $50,000",
"$50,000 under $75,000"),
"75K - <200K" = c("$75,000 under $100,000",
"$100,000 under $200,000"
),
"200K+" = c("$200,000 under $500,000",
"$500,000 under $1,000,000",
"$1,000,000 under $1,500,000",
"$1,500,000 under $2,000,000",
"$2,000,000 under $5,000,000",
"$5,000,000 under $10,000,000",
"$10,000,000 or more")
))
# combine the income levels to smaller groups
df_filter_sum <- df_filter %>%
mutate(Value = gsub(",", "", Value), # Remove commas
Value = as.numeric(Value)) # Convert to numeric
df_filter_sum <- df_filter_sum %>%
group_by(adjusted_gross_income) %>%
summarise(
sum_return_thous = sum(Value)/1e3,
)

df_filter_sum<- df_filter_sum %>%
mutate(adjusted_gross_income = 
      factor(adjusted_gross_income, levels = 
      c("<25K", "25K - <75K","75K - <200K", "200K+")))
                                                                        # organize the bracket to smallest to largest
                                      
                                      
ggplot(df_filter_sum, aes(x = adjusted_gross_income, y = sum_return_thous)) +
geom_col(fill = "steelblue") +
labs(title = "Count of Tax Returns Filed by Income Group",
x = "Income Group",
y = "Count (by thousands)")
```
I filter the type of returns here to get the total returns filed by income group. In this dataset, there is a
difference between all returns and taxable returns so in this case, we are looking at all returns. Since the
source data had an overwhelming amount of income groups, I combined them to offer a more wider view of
the analysis.

# Effective tax rate
```{r }

df_filter_income <- df_tidy %>%
filter(Return_Type == "taxable",
Category %in% c("total_income_tax", "taxable_income"),
Metric %in% c("amount"))

df_filter_income <- df_filter_income %>%
mutate(adjusted_gross_income = fct_collapse(adjusted_gross_income,
"<25K" = c("No adjusted gross income",
"$1 under $5,000",
"$5,000 under $10,000",
"$10,000 under $15,000",
"$15,000 under $20,000",
"$20,000 under $25,000"),
"25K - <75K" = c("$25,000 under $30,000",
"$30,000 under $40,000",
"$40,000 under $50,000",
"$50,000 under $75,000"),
"75K - <200K" = c("$75,000 under $100,000",
"$100,000 under $200,000"
),
"200K+" = c("$200,000 under $500,000",
"$500,000 under $1,000,000",
"$1,000,000 under $1,500,000",
"$1,500,000 under $2,000,000",
"$2,000,000 under $5,000,000",
"$5,000,000 under $10,000,000",
"$10,000,000 or more")
))

# combine the income levels to smaller groups
df_filter_incometax <- df_filter_income %>%
mutate(Value = gsub(",", "", Value),
Value = as.numeric(Value))
df_avg_taxes <- df_filter_incometax %>%
group_by(adjusted_gross_income) %>%
summarise(avg_paid_taxes = sum(Value[Category == "total_income_tax"], na.rm = TRUE) /
sum(Value[Category == "taxable_income"], na.rm = TRUE))

df_avg_taxes<- df_avg_taxes %>%
mutate(adjusted_gross_income = factor(adjusted_gross_income, levels = c("<25K", "25K - <75K","75K - <200K","200K+")))
                                                                        
                                                                        # organize the bracket to smallest to largest
ggplot(df_avg_taxes, aes(x = adjusted_gross_income, y = avg_paid_taxes)) +
geom_col(fill = "steelblue") +
labs(title = "Effective tax rate by Income Group",
x = "Income Group",
y = "Effective tax rate")

```
Same thing I did here with the total count of returns. The only difference here is that I also average to get
the effective tax rate, which is the total income tax/total count of taxable income returns.

# Conclusion


After transforming the file to a structured table that we can use for analysis, I aggregated and grouped the
income categories to generate insights. One key observation is that the majority of tax filers fall within the
$25K to under $75K income bracket, as indicated by the total number of returns filed for the 2022 tax year.
Additionally, I calculated the effective tax rate for each income group by dividing the total income tax paid
by the total number of returns, allowing for a comparative view of tax burdens across different income levels.
For future analysis, it would be valuable to compare tax years, particularly during periods of major tax
reform, such the Tax Cuts and Jobs Act (2018–2025). Examining changes in tax rates, income distribution,
and overall tax burdens over time could provide deeper insights into the impact of tax policy.
Take Away from the Assignment- The janitor package, particularly the clean_names() function, proved to
be extremely useful. It ensures consistent formatting of column names, which is essential when applying
pattern-based transformations or reshaping the data into a long format.