---
title: 'DATA 607: Week 3 - Resubmission'
author: "Cindy Lin"
date: "2025-2-13"
output:
  pdf_document: default
  html_document: default
---


# INTRODUCTION
There are 4 parts of the assignment with the goal to showcase database normalization and manioulating data in R. 

## Loading library

```{r library, message= FALSE}
library(dplyr)

```
I am loading the necessary library to perform the tasks assigned 

## Part I - Normalization
```{r part I, message=FALSE}


students <- data.frame(
  student_id = c(1,2,3,4),
  name = c("Spongebob", "Sandy", "Patrick", "Squidward"),
  class_level = c("Freshman", "Sophomore", "Junior", "Senior")
)

courses <- data.frame (
  course_id = c(001, 002, 003, 004),
  course_name = c("English 101", "English 102", "Math 101", "PHIL 101")
)


grades <- data.frame (
  grade_id = c(1,2,3,4,5,6),
  grade = c("A", "A-", "B", "C", "D", "F")
)


Enrollment <- data.frame(
  grade_id = c(1,2,3,4,5,6),
  student_id = c(1,2,3,4,1,2),
  course_id = c(001, 002, 003, 004, 001, 002)
)


print(students)
print(courses)
print(Enrollment)
```
Goal of normalization - efficiency, consistency, and accuracy. 
Example - 
Students, courses, grades, and enrollment tables show 1NF as it has a primary key, consistent data type and no repeating groups in the column. 

Students, courses, grades, and enrollment tables shows 2NF as the attributes are dependent on the primary key (enrollment table is dependent on composite key).

Students, courses, grades, and enrollment tables shows 3NF as no other non-key attributes are dependent on other non-key attributes (enrollment table is dependent on both composite key - student ID and course ID).



## Part II - Character Manipulation

```{r part II, message=FALSE}

get_data <- "https://raw.githubusercontent.com/fivethirtyeight/data/refs/heads/master/college-majors/majors-list.csv"

majors <- data.frame(read.csv(get_data))

selected_majors <- grep("DATA|STATISTICS", majors$Major, value = TRUE, ignore.case = TRUE)

# Print the results
print(selected_majors)



```
Loading the data from github and using the function "grep()" to search "Data" and "Statistics" from the list of college majors. 


## Part III

```{r part III, message=FALSE}

strings <- c("aaabbb", "xyyx", "abba", "abc", "xyz", "hello", "wow", "123", "abab", "abaca", "abcdcba", "wwwee", "ww", "ana", "awaw", "axabaea", "zywvvwyz")

matches <- grep("(.)\\1\\1", strings, value = TRUE)
matches1 <- grep("(.)(.)\\2\\1", strings, value = TRUE)
matches2 <- grep("(..)\\1", strings, value = TRUE)


matches3 <- grep("(.).\\1.\\1", strings, value = TRUE)
#find matches that have the same beginning character and have that repeat every other character

matches4 <- grep("(.)(.)(.).*\\3\\2\\1", strings, value = TRUE)
#find matches that have at leat 3 chracter that repeats in mirror order. 


print(matches)
print(matches1)
print(matches2)
print(matches3)
print(matches4)

```
(.)\1\1
Explain - 
(.) - capture a group 
\1  - matches the character in the capture (above)
\1  - repeat above again

what the code does- 
find matches that have at least 3 of the same repeating character

"(.)(.)\\2\\1"
Explain - 
(.) - capture the first group
(.) - capture the second group
\\2 - match the character in the second capture
\1  - match the character in the first capture

what the code does- 
find matches that have that first and last character to be the same and have at least one the same repeating character in between. I think of it as a mirror pattern. 

(..)\1
Explain - 
(..) - capture a group that matches two character
\1  - matches the character in the capture (above)

what the code does- 
find matches that have alternating repeating characters

"(.).\\1.\\1"
Explain - 
(.) - capture a group 
.   - 
\1  - matches the character in the capture (above)
\1  - repeat above again

what the code does- 
find matches that have at least 3 of the same repeating character
"(.)(.)(.).*\\3\\2\\1"

## Part IV

```{r part IV, message=FALSE}

stringstwo <- c("rar", "hello", "papa", "aba", "aweerrttffa", "chyech", "eleven")

matches5 <- grep("^(.).*\\1$", stringstwo, value = TRUE)
#Start and end with the same character.

matches6 <- grep("(..).*\\1", stringstwo, value = TRUE)
#Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)

matches7 <- grep("(.).*\\1.*\\1", stringstwo, value = TRUE)
#Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)


print(matches5)
print(matches6)
print(matches7)
```

# CONCLUSION 
In this week's assignment, we explored various data manipulation techniques, which includes normalization, character manipulation, and regular expression. These are important for organizing and analyzing with greater efficiency. 