---
title: 'DATA 607: Week 9'
author: "Cindy Lin"
date: "2025-03-26"
output:
  pdf_document: default
  html_document: default
  
---

# INTRODUCTION

Week 9 assignment's goal is to pull information from one of the New York Times API and read it in the JSON data, and transform it into an R Data Frame.


## Loading library

```{r library, include= FALSE}

install.packages("httr", repos = "https://cloud.r-project.org")
options(repos = c(CRAN = "https://cloud.r-project.org/"))


install.packages("jsonlite")
install.packages("dplyr") 

```
installing the necessary packages for API and JSON data



```{r API}

library(httr)
library(jsonlite)
library(dplyr)
library(purrr)


api_key <- Sys.getenv("API")
url <- "https://api.nytimes.com/svc/books/v3/lists/best-sellers/history.json"

fetch_page <- function(offset = 0) {
  response <- GET(url, query = list(
    `api-key` = api_key,
    offset = offset
  ))
  result <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
  
  return(result$results)
}

offsets <- seq(0, 100, by = 20)  
all_books <- map_dfr(offsets, fetch_page)

head(all_books)

```
The default for the NYT API is set to limit to fetch 20 but I wanted to have more results so utilized "offset" to get 100 rows by dataset. 


```{r analysis}

library(dplyr)

top_authors <- all_books %>%
  group_by(author) %>%
  summarise(book_count = n()) %>%
  arrange(desc(book_count))

head(top_authors, 5)


top_publishers <- all_books %>%
  group_by(publisher) %>%
  summarise(book_count = n()) %>%
  arrange(desc(book_count))

head(top_publishers, 5)


```
This analyze the top 5 author and the top 5 publisher in the data set. 

# CONCLUSION
This was a straight forward approach to acquiring the data through an API. As long as I can attain the API key, I was able to pull the data into a data frame. 

