---
title: 'DATA 607: Week 6.3'
author: "Cindy Lin and William Forero"
date: "2025-03-05"
output:
  pdf_document: default
  html_document: default
  
---

# INTRODUCTION

Week 6 project's goal is to transform three untidy data set and tidy/transform them and generate the discussed analysis. 

Data Set#3 - Sales data


## Loading library

```{r library, message= FALSE}

library (tidyverse)

```
I am loading the tidyverse library because there are functions that can help tidy the loaded data. 

## Loading the data and removing rows
```{r getdata, message=FALSE}

get_data <- read.csv("Sales.csv")
# Skip the first row to get the header

glimpse(get_data)
```
Loading the data 


## Long format
```{r long format data}

df_long <- get_data %>%
  pivot_longer(
    cols = starts_with("Jan") | starts_with("Feb") | starts_with("Mar") | 
           starts_with("Apr") | starts_with("May") | starts_with("Jun"),
    names_to = "Month",
    values_to = "Sales"
  ) %>%
  mutate(Month = gsub(".Sales", "", Month))  #

df_long
```
Shaping to long format for analysis


# Year to Year Trend
```{r }

df_long$Month <- factor(df_long$Month, levels = 
                        c("Jan", "Feb", "Mar", "Apr", "May", "Jun"))

sales_trends <- df_long %>%
  group_by(Product.Name, Month) %>%
  summarise(Total_Sales = sum(Sales))
#sum of sales by product

sales_trends_region <- df_long %>%
  group_by(Region, Month) %>%
  summarise(Total_Sales = sum(Sales))
#sum of sales by product

ggplot(sales_trends, aes(x = Month, 
                         y = Total_Sales, 
                         group = Product.Name , 
                         color = Product.Name)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(title = "Year-over-Year Sales by Product", x = "Month", y = "Total Sales") +
  theme_minimal()
# plot trends

ggplot(sales_trends_region, aes(x = Month, y = Total_Sales, group = Region , color = Region)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(title = "Year-over-Year Sales by Region", x = "Month", y = "Total Sales") +
  theme_minimal()
# plot trends
```
June sales tends to have a higher for all products. It seems that for the most part, sales increase throughout the year. With Product B having the most sales and Prodcut C with the less amount of the three. 

The East region also have the higher sales with north having less than the three.

# Product Sales Distribution
```{r }
sales_trends2 <- df_long %>%
  group_by(Product.Name) %>%
  summarise(Total_Sales = sum(Sales))

df_long |>
  group_by(Product.Name) |>
  summarise(Total_Sales = sum(Sales, na.rm = TRUE),
            Average_Sales = mean(Sales),
            Max_Sales = max(Sales),
            Min_Sales = min(Sales),
            .groups = "drop") 

ggplot(sales_trends2, aes(x = Product.Name, 
                          y = Total_Sales, 
                          fill = Product.Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Product Sales Distribution", x = "Product", y = "Total Sales") +
  theme_minimal()

```
Product B had roughly 5000 total sales, Product A has a little over 4000 sales, and Product C has 2000 total sales.

# Regional Performance Comparison
```{r }

sales_trends_region2 <- df_long %>%
  group_by(Region) %>%
  summarise(Total_Sales = sum(Sales))
#sum of sales by product

df_long |>
  group_by(Region) |>
  summarise(Total_Sales = sum(Sales, na.rm = TRUE),
            Average_Sales = mean(Sales),
            Max_Sales = max(Sales),
            Min_Sales = min(Sales),
            .groups = "drop")

ggplot(sales_trends_region2, aes(x = Region, 
                                 y = Total_Sales, 
                                 fill = Region)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Regional Sales Performance", x = "Region", y = "Total Sales") +
  theme_minimal()

```
East region has roughly over 5000 sales total, north has a little over 2000 sales, and South has around 3500 total sales.


# By Product and Region
```{r }

df_long |>
  group_by(Region, Product.Name) |>
  summarise(Total_Sales = sum(Sales, na.rm = TRUE),
            Average_Sales = mean(Sales),
            Max_Sales = max(Sales),
            Min_Sales = min(Sales),
            .groups = "drop") 

ggplot(df_long, aes(x=Region, y= Sales, color = Product.Name)) + geom_point(postion = "jitter")
```
Here we see that product A and B are relately close in sales for all 3 regions but for product C, the difference between the other two is greater. We can see the largest difference in East region. 


# Conclusion
Visualizations provide us quick insights on the data we are looking at, and in this scenario, it was helpfully in concluding the results of the sales and which regions are performing better than the rest. In this case, we see the the East region and product B performed the best versus the north and product C performed the worse. 
